---
title: "Clinical Outcome Prediction"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Clinical Outcome Prediction}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette demonstrates how MOSAIC's connectivity-derived features can be used for **clinical outcome prediction**, providing a complementary signal to conventional abundance-based analysis. We apply MOSAIC to a large-scale COVID-19 scRNA-seq cohort and show that integrating connectivity and abundance features improves disease severity classification in monocytes.

## Dataset

We use a publicly available COVID-19 scRNA-seq dataset ([Su et al., 2020](https://doi.org/10.1016/j.cell.2020.10.037)) comprising peripheral blood mononuclear cells (PBMCs) from 151 COVID-19 patients stratified by disease severity (Moderate vs. Severe). Control samples are excluded. We focus on **monocytes**, which were identified as the strongest immune driver of severity in our unbiased screen across cell lineages.

## Data Preprocessing

```{r load-data}
library(MOSAIC)
library(Seurat)
library(schard)
library(glmnet)
library(pROC)
library(caret)
library(ggplot2)
library(dplyr)

# Load the COVID-19 scRNA-seq data
seurat_obj <- schard::h5ad2seurat("path/to/Covid_19_PBMC.h5ad", use.raw = TRUE)

# Subset to monocytes, exclude controls
seurat_mono <- seurat_obj[, seurat_obj$majorType == "Mono"]
seurat_mono <- seurat_mono[, seurat_mono$`CoVID-19 severity` != "control"]

# Remove samples with too few cells
sample_counts <- table(seurat_mono$sampleID)
samples_to_remove <- names(sample_counts[sample_counts < 10])
seurat_mono <- subset(seurat_mono, subset = sampleID %ni% samples_to_remove)
```

### Feature Selection

```{r feature-selection}
counts <- GetAssayData(seurat_mono, assay = "RNA", slot = "counts")
sample_ids <- unique(as.character(seurat_mono$sampleID))

genes_per_sample <- lapply(sample_ids, function(id) {
  cells <- which(seurat_mono$sampleID == id)
  rownames(counts)[rowSums(counts[, cells, drop = FALSE]) > 3]
})
genes_to_keep <- Reduce(intersect, genes_per_sample)

seurat_rna <- subset(seurat_mono, features = genes_to_keep)
seurat_rna <- NormalizeData(seurat_rna)
```

## Step 1: Run MOSAIC (Single Modality)

MOSAIC's mathematical framework is modality-agnostic. Even with a single modality (RNA), it captures latent regulatory information from the feature connectivity structure.

```{r run-mosaic}
result <- run_MOSAIC(
  list(RNA = seurat_rna),
  assays = c("RNA"),
  sample_meta = "sampleID",
  condition_meta = "CoVID-19 severity"
)
```

## Step 2: Prepare Prediction Inputs

We construct two types of patient-level feature vectors:

1. **Connectivity features (MOSAIC)**: Each patient's embedding matrix is vectorized into a single feature vector.
2. **Abundance features (Pseudobulk)**: Raw counts are aggregated per patient and log-normalized.

```{r prepare-inputs}
# --- Connectivity features ---
mosaic_vectors <- lapply(result$mosaic_embed_list, as.vector)
X_mosaic <- do.call(rbind, mosaic_vectors)
label <- result$annotation$Condition
y <- ifelse(label %in% c("severe/critical"), 1, 0)

# --- Abundance features (pseudobulk) ---
pseudobulk_list <- AggregateExpression(
  seurat_rna,
  group.by = "sampleID",
  assays = "RNA",
  slot = "counts",
  normalization.method = NULL
)
pseudobulk_counts <- pseudobulk_list$RNA
pb_obj <- CreateSeuratObject(counts = pseudobulk_counts)
pb_obj <- NormalizeData(pb_obj, normalization.method = "LogNormalize", scale.factor = 10000)
X_expr <- t(as.matrix(GetAssayData(pb_obj, slot = "data")))

# --- Align sample order ---
common_samples <- intersect(rownames(X_mosaic), rownames(X_expr))
X_mosaic <- X_mosaic[common_samples, , drop = FALSE]
X_expr <- X_expr[common_samples, , drop = FALSE]

y_map <- setNames(label, rownames(result$annotation))
y <- ifelse(y_map[common_samples] %in% c("severe/critical"), 1, 0)
```

## Step 3: Repeated Stratified Cross-Validation

We train L1-regularized logistic regression (Lasso) classifiers using three inputs: MOSAIC connectivity features alone, pseudobulk abundance features alone, and a late-fusion integrated model.

```{r cross-validation}
n_repeats <- 10
n_folds <- 5

results_mosaic <- c()
results_expr <- c()
results_comb <- c()

all_labels <- c()
all_preds_mos <- c()
all_preds_exp <- c()
all_preds_comb <- c()

round_exp_list <- list()
round_mos_list <- list()

for (seed in seq_len(n_repeats)) {
  set.seed(seed)
  folds <- createFolds(factor(y), k = n_folds, list = FALSE)

  round_mos <- rep(NA, length(y))
  round_exp <- rep(NA, length(y))

  for (i in seq_len(n_folds)) {
    train_idx <- which(folds != i)
    test_idx <- which(folds == i)

    # MOSAIC (Connectivity)
    set.seed(seed * 1000 + i)
    cv_mos <- cv.glmnet(X_mosaic[train_idx, ], y[train_idx],
                         family = "binomial", alpha = 1, standardize = TRUE)
    round_mos[test_idx] <- predict(cv_mos, X_mosaic[test_idx, ],
                                    s = "lambda.min", type = "response")

    # Pseudobulk (Abundance)
    set.seed(seed * 1000 + i)
    cv_exp <- cv.glmnet(X_expr[train_idx, ], y[train_idx],
                         family = "binomial", alpha = 1, standardize = TRUE)
    round_exp[test_idx] <- predict(cv_exp, X_expr[test_idx, ],
                                    s = "lambda.min", type = "response")
  }

  auc_mos <- auc(roc(y, as.numeric(round_mos), quiet = TRUE))
  auc_exp <- auc(roc(y, as.numeric(round_exp), quiet = TRUE))
  round_comb <- (round_mos + round_exp) / 2
  auc_comb <- auc(roc(y, as.numeric(round_comb), quiet = TRUE))

  results_mosaic <- c(results_mosaic, auc_mos)
  results_expr <- c(results_expr, auc_exp)
  results_comb <- c(results_comb, auc_comb)

  all_labels <- c(all_labels, y)
  all_preds_mos <- c(all_preds_mos, round_mos)
  all_preds_exp <- c(all_preds_exp, round_exp)
  all_preds_comb <- c(all_preds_comb, round_comb)

  round_exp_list <- append(round_exp_list, list(round_exp))
  round_mos_list <- append(round_mos_list, list(round_mos))
}

cat("Mean MOSAIC AUC:", round(mean(results_mosaic), 3), "\n")
cat("Mean Expression AUC:", round(mean(results_expr), 3), "\n")
cat("Mean Combined AUC:", round(mean(results_comb), 3), "\n")
```

## Figure 6B: ROC Curves

```{r fig6b-roc, fig.width = 5.4, fig.height = 4.6}
roc_exp_full <- roc(all_labels, all_preds_exp)
roc_mos_full <- roc(all_labels, all_preds_mos)
roc_comb_full <- roc(all_labels, all_preds_comb)

roc_data <- rbind(
  data.frame(FPR = 1 - roc_exp_full$specificities,
             TPR = roc_exp_full$sensitivities,
             Model = "Pseudobulk (Abundance)"),
  data.frame(FPR = 1 - roc_mos_full$specificities,
             TPR = roc_mos_full$sensitivities,
             Model = "MOSAIC (Connectivity)"),
  data.frame(FPR = 1 - roc_comb_full$specificities,
             TPR = roc_comb_full$sensitivities,
             Model = "Integrated (Connectivity + Abundance)")
)
roc_data$Model <- factor(roc_data$Model, levels = c(
  "Integrated (Connectivity + Abundance)",
  "MOSAIC (Connectivity)",
  "Pseudobulk (Abundance)"
))

stats_text <- paste0(
  "AUC Scores\n",
  "Integrated:  ", sprintf("%.3f", mean(results_comb)), "\n",
  "MOSAIC:      ", sprintf("%.3f", mean(results_mosaic)), "\n",
  "Pseudobulk:  ", sprintf("%.3f", mean(results_expr))
)

ggplot(roc_data, aes(x = FPR, y = TPR, color = Model, linewidth = Model)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray80") +
  geom_line() +
  scale_color_manual(values = c("#56B4E9", "#E69F00", "grey60")) +
  scale_linewidth_manual(values = c(1.5, 1.2, 0.8)) +
  annotate("text", x = 1, y = 0.35, label = stats_text,
           hjust = 1, vjust = 0, size = 2.8, lineheight = 1.0) +
  labs(title = "Predictive Performance for Monocytes",
       x = "False Positive Rate", y = "True Positive Rate") +
  theme_classic(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = c(1, 0.0),
    legend.justification = c("right", "bottom"),
    legend.background = element_blank(),
    legend.title = element_blank(),
    legend.key.width = unit(1.0, "cm"),
    legend.text = element_text(size = 10)
  )
```

## Figure 6C: Complementary Risk Stratification

We compare patient-specific risk scores from the Abundance (pseudobulk) and Connectivity (MOSAIC) models, revealing that the two modalities capture orthogonal dimensions of disease severity.

```{r fig6c-scatter, fig.width = 8, fig.height = 6}
plot_data <- data.frame(
  Expr_Risk = as.numeric(rowMeans(do.call(cbind, round_exp_list))),
  Mos_Risk = as.numeric(rowMeans(do.call(cbind, round_mos_list))),
  Outcome = ifelse(y == 1, "Severe", "Moderate")
)

mos_exp_cor <- cor(plot_data$Expr_Risk, plot_data$Mos_Risk)

# Define clinical categories
plot_data$Category <- "Consensus"
plot_data$Category[plot_data$Expr_Risk < 0.5 & plot_data$Mos_Risk > 0.5 &
                     plot_data$Outcome == "Severe"] <- "Type_II_Rescue"
plot_data$Category[plot_data$Expr_Risk > 0.5 & plot_data$Mos_Risk < 0.5 &
                     plot_data$Outcome == "Moderate"] <- "Specificity_Win"
plot_data$Category[plot_data$Expr_Risk > 0.5 & plot_data$Mos_Risk < 0.5 &
                     plot_data$Outcome == "Severe"] <- "Type_I_Rescue"

plot_data$Category <- factor(plot_data$Category,
  levels = c("Consensus", "Type_II_Rescue", "Type_I_Rescue", "Specificity_Win"),
  labels = c("Consensus Severe", "Connectivity-Rescued",
             "Abundance-Rescued", "Abundance False Positive")
)

color_map <- c("Moderate" = "#56B4E9", "Severe" = "#D55E00")

ggplot(plot_data, aes(x = Expr_Risk, y = Mos_Risk)) +
  annotate("rect", xmin = 0, xmax = 0.5, ymin = 0.5, ymax = 1.05,
           fill = "#FFFFE0", alpha = 0.3) +
  annotate("rect", xmin = 0.5, xmax = 1.05, ymin = 0, ymax = 0.5,
           fill = "#E6E6FA", alpha = 0.3) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "grey60") +
  geom_vline(xintercept = 0.5, linetype = "dashed", color = "grey60") +
  geom_point(data = subset(plot_data, Category == "Consensus Severe"),
             aes(color = Outcome, shape = Category), size = 3, alpha = 0.6) +
  geom_point(data = subset(plot_data, Category != "Consensus Severe"),
             aes(color = Outcome, shape = Category), size = 3.5, stroke = 1.2) +
  annotate("text", x = 0.05, y = 0.95, label = "Type II: Connectivity-High",
           hjust = 0, fontface = "bold", color = "#D55E00", size = 3.5) +
  annotate("text", x = 0.95, y = 0.10, label = "Type I: Abundance-High",
           hjust = 1, fontface = "bold", color = "darkblue", size = 3.5) +
  scale_color_manual(values = color_map) +
  scale_shape_manual(values = c("Connectivity-Rescued" = 17,
                                "Consensus Severe" = 19,
                                "Abundance-Rescued" = 15,
                                "Abundance False Positive" = 1)) +
  labs(
    title = "Complementary Mechanisms of Severity",
    subtitle = paste0("Pearson R = ", round(mos_exp_cor, 2)),
    x = "Abundance Risk Score (Pseudobulk)",
    y = "Connectivity Risk Score (MOSAIC)",
    color = "Clinical Outcome",
    shape = "Patient Subtype"
  ) +
  theme_bw(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5, color = "grey30", size = 11),
    legend.position = "right",
    panel.grid.minor = element_blank()
  )
```

## Figure 6E: Top Predictive Features

We extract the Lasso-selected features from each model and rank them by absolute coefficient magnitude to identify the top molecular drivers.

```{r fig6e-drivers, fig.width = 10, fig.height = 5}
library(cowplot)

# --- Fit final models on full data ---
fit_expr <- glmnet(X_expr, y, family = "binomial", alpha = 1, standardize = TRUE)
coef_expr <- coef(fit_expr, s = cv_exp$lambda.min)

fit_mos <- glmnet(X_mosaic, y, family = "binomial", alpha = 1, standardize = TRUE)
coef_mos <- coef(fit_mos, s = cv_mos$lambda.min)

# --- Abundance drivers ---
df_expr <- data.frame(
  Feature = rownames(coef_expr),
  Weight = as.numeric(coef_expr)
) %>%
  filter(Feature != "(Intercept)", Weight != 0) %>%
  mutate(AbsWeight = abs(Weight)) %>%
  arrange(desc(AbsWeight)) %>%
  slice_head(n = 15) %>%
  mutate(RelWeight = (AbsWeight / max(AbsWeight)) * 100)

df_expr$Feature <- factor(df_expr$Feature, levels = df_expr$Feature[order(df_expr$RelWeight)])

# --- Connectivity drivers (aggregate across latent dimensions) ---
feature_names <- rownames(result$mosaic_embed_list[[1]])
dim_r <- ncol(result$mosaic_embed_list[[1]])

df_mos_raw <- data.frame(
  Feature = c("(Intercept)", rep(feature_names, dim_r)),
  Weight = as.numeric(coef_mos)
) %>%
  filter(Feature != "(Intercept)", Weight != 0)

df_mos_agg <- df_mos_raw %>%
  group_by(Feature) %>%
  summarise(AbsWeight = sum(abs(Weight))) %>%
  arrange(desc(AbsWeight)) %>%
  slice_head(n = 15) %>%
  mutate(RelWeight = (AbsWeight / max(AbsWeight)) * 100)

df_mos_agg$Feature <- factor(df_mos_agg$Feature,
                              levels = df_mos_agg$Feature[order(df_mos_agg$RelWeight)])

# --- Lollipop plot function ---
plot_drivers <- function(data, title_text, point_color) {
  ggplot(data, aes(x = RelWeight, y = Feature)) +
    geom_segment(aes(x = 0, xend = RelWeight, y = Feature, yend = Feature),
                 color = "grey70") +
    geom_point(color = point_color, size = 4) +
    theme_bw(base_size = 14) +
    labs(title = title_text, x = "Relative Importance (%)", y = "") +
    theme(
      axis.text.y = element_text(face = "bold", color = "black"),
      panel.grid.minor = element_blank(),
      panel.border = element_blank(),
      axis.line.x = element_line(color = "black"),
      axis.ticks.y = element_blank(),
      plot.title = element_text(hjust = 0.5, face = "bold", size = 12),
      legend.position = "none"
    )
}

p_left <- plot_drivers(df_expr, "Top Abundance Drivers", "grey40")
p_right <- plot_drivers(df_mos_agg, "Top Connectivity Drivers", "#E69F00")

plot_grid(p_left, p_right, ncol = 2, align = "h")
```

The minimal overlap between the two sets of predictive features confirms that connectivity and abundance capture complementary dimensions of disease variation.

## Session Info

```{r session-info}
sessionInfo()
```
