---
title: "Differential Connectivity Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Differential Connectivity Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette demonstrates MOSAIC's **Differential Connectivity (DC) analysis** pipeline. DC analysis identifies features whose regulatory network partners are rewired across conditions, even when their expression levels remain unchanged. We apply MOSAIC to a multi-omic CITE-seq dataset of human T cells profiled before and after vaccination.

## Dataset

We use a publicly available CITE-seq dataset ([Hao et al., 2021](https://doi.org/10.1016/j.cell.2021.04.048)) containing paired RNA and protein (ADT) measurements from human PBMCs across 8 donors, profiled at Day 0 (naive) and Day 7 (activated) post-vaccination. We focus on **CD4 Naive T cells** to isolate activation-specific signals.

## Data Preprocessing

```{r load-data}
library(MOSAIC)
library(Seurat)
library(SeuratDisk)
library(ggplot2)
library(dplyr)

# Load the CITE-seq data
pbmc <- LoadH5Seurat("/banach2/chang/cleaned_rong_project/dataset/multi.h5seurat")

# Subset to CD4 Naive T cells, Day 0 vs Day 7
seurat_subset <- subset(pbmc, idents = "CD4 Naive")
seurat_subset <- seurat_subset[, seurat_subset$time %in% c(0, 7)]
```

### Feature Selection

We retain features expressed (total count > 3) in every sample, ensuring all features are measurable across the cohort.

```{r feature-selection}
# RNA features
counts_rna <- GetAssayData(seurat_subset, assay = "SCT", slot = "counts")
sample_ids <- unique(as.character(seurat_subset$orig.ident))

genes_per_sample <- lapply(sample_ids, function(id) {
  cells <- which(seurat_subset$orig.ident == id)
  rownames(counts_rna)[rowSums(counts_rna[, cells, drop = FALSE]) > 3]
})
genes_to_keep <- Reduce(intersect, genes_per_sample)

# ADT features
counts_adt <- GetAssayData(seurat_subset, assay = "ADT", slot = "counts")
proteins_per_sample <- lapply(sample_ids, function(id) {
  cells <- which(seurat_subset$orig.ident == id)
  rownames(counts_adt)[rowSums(counts_adt[, cells, drop = FALSE]) > 3]
})
proteins_to_keep <- Reduce(intersect, proteins_per_sample)

# Subset features
seurat_subset[["SCT"]] <- subset(seurat_subset[["SCT"]], features = genes_to_keep)
seurat_subset[["ADT"]] <- subset(seurat_subset[["ADT"]], features = proteins_to_keep)
```

### Prepare Input for MOSAIC

MOSAIC takes a list of normalized Seurat objects, one per modality. We split the combined object and normalize each modality separately.

```{r prepare-input}
# Add sample and condition metadata
seurat_subset$sample_id <- seurat_subset$orig.ident
seurat_subset$condition <- seurat_subset$time

# Create feature name vector for later use
feature_names <- c(
  paste0("RNA_", rownames(seurat_subset[["SCT"]])),
  paste0("ADT_", rownames(seurat_subset[["ADT"]]))
)

# Separate into per-modality Seurat objects
seurat_rna <- seurat_subset
seurat_rna[["ADT"]] <- NULL

seurat_adt <- seurat_subset
DefaultAssay(seurat_adt) <- "ADT"
seurat_adt[["SCT"]] <- NULL

# Normalize
seurat_rna <- NormalizeData(seurat_rna)
seurat_adt <- NormalizeData(seurat_adt, normalization.method = "CLR")
```

## Step 1: Run MOSAIC Embedding

```{r run-mosaic}
result <- run_MOSAIC(
  seurat_list = list(RNA = seurat_rna, ADT = seurat_adt),
  assays = c("SCT", "ADT"),
  sample_meta = "sample_id",
  condition_meta = "condition"
)
```

### Visualize Eigenvalue Spectrum

The eigenvalue spectrum from the aggregated spectral decomposition reveals the effective dimensionality of the shared latent space. The red dashed line marks the elbow point automatically selected by the Kneedle algorithm.

```{r eigenplot, fig.width = 5, fig.height = 3.5}
elbow <- find_elbow_kneedle(result$eigenvalues)

plot_eigen(result$eigenvalues) +
  geom_vline(xintercept = elbow, color = "red", linetype = "dashed") +
  ggtitle(paste0("Global eigenvalues (elbow at ", elbow, ")"))
```

## Step 2: Run DC Test

We run the per-feature PERMANOVA test on the real data to quantify how much each feature's embedding shifts between Day 0 and Day 7.

```{r dc-test}
n_sample <- length(result$mosaic_embed_list)

dc_result <- run_DC_test(
  result$mosaic_embed_list,
  n_sample = n_sample,
  groups = result$annotation$Condition,
  dist_method = "euclidean"
)
```

## Step 3: Construct Null Distribution

To calibrate statistical significance, we shuffle the cell-to-sample labels to destroy condition-specific signals, re-run MOSAIC, and collect null F-statistics.

```{r shuffle-null}
# Shuffle sample labels
set.seed(123)
shuffled_indices <- sample(nrow(seurat_rna@meta.data))
seurat_rna$shuffled_sample_id <- seurat_rna$sample_id[shuffled_indices]
seurat_rna$shuffled_condition <- seurat_rna$condition[shuffled_indices]
seurat_adt$shuffled_sample_id <- seurat_adt$sample_id[shuffled_indices]
seurat_adt$shuffled_condition <- seurat_adt$condition[shuffled_indices]

# Run MOSAIC on shuffled data
shuffle_result <- run_MOSAIC(
  seurat_list = list(RNA = seurat_rna, ADT = seurat_adt),
  assays = c("SCT", "ADT"),
  sample_meta = "shuffled_sample_id",
  condition_meta = "shuffled_condition",
  verbose = FALSE
)

# Run DC test on shuffled data
shuffle_dc <- run_DC_test(
  shuffle_result$mosaic_embed_list,
  n_sample = n_sample,
  groups = shuffle_result$annotation$Condition,
  dist_method = "euclidean"
)
```

## Step 4: Compute Empirical P-Values and Identify DC Features

We compare each feature's observed F-statistic against the pooled null distribution of F-statistics from the shuffled run.

```{r empirical-pvalues}
F_obs <- unlist(dc_result$F_stats_list)
F_null <- unlist(shuffle_dc$F_stats_list)

pvalues <- sapply(F_obs, function(x) calculate_empirical_pvalue(x, F_null))

# Identify significant DC features
dc_idx <- which(pvalues < 0.05)
cat(length(dc_idx), "DC features identified out of", length(pvalues), "total features\n")
```

### DC Feature Table

```{r dc-table}
dc_table <- data.frame(
  Feature = feature_names[dc_idx],
  F_statistic = F_obs[dc_idx],
  R2 = dc_result$r2_list[dc_idx],
  Empirical_pvalue = pvalues[dc_idx]
)
dc_table <- dc_table[order(dc_table$Empirical_pvalue, -dc_table$F_statistic), ]

head(dc_table, 20)
```

## Step 5: Pathway Enrichment Analysis

We perform Reactome pathway enrichment on the DC features to assess their biological coherence.

```{r pathway-analysis}
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)

# Extract gene symbols from DC features
dc_feature_names <- feature_names[dc_idx]
dc_gene_symbols <- sub("^(RNA_|ADT_)", "", dc_feature_names)

# Convert to Entrez IDs
entrez_ids <- bitr(dc_gene_symbols,
                   fromType = "SYMBOL",
                   toType = "ENTREZID",
                   OrgDb = org.Hs.eg.db)

# Reactome pathway enrichment
reactome_results <- enrichPathway(
  gene = entrez_ids$ENTREZID,
  pvalueCutoff = 0.05,
  readable = TRUE
)

dotplot(reactome_results, showCategory = 15, font.size = 8) +
  ggtitle("Reactome Pathway Enrichment of DC Features")
```

## Step 6: Visualize a DC Feature

We select a top DC feature and visualize its sample-level connectivity profile using MDS. Each point represents one sample's embedding of this feature, colored by timepoint.

```{r mds-plot, fig.width = 5, fig.height = 4}
# Select a top DC feature (e.g., STAT5B)
chosen_feature <- "RNA_STAT5B"
chosen_idx <- which(feature_names == chosen_feature)

# Get the per-feature similarity matrix from DC results
sim_mat <- dc_result$similarity_matrix_list[[chosen_idx]]
groups <- dc_result$group_list[[chosen_idx]]

# MDS visualization
plot_mds_cluster(
  sim_mat,
  title = paste(chosen_feature, "Connectivity Profile"),
  cluster = paste0("Day_", groups),
  custom_colors = c("Day_0" = "#56B4E9", "Day_7" = "#E69F00"),
  size = 2,
  ellipse_level = 0.68
)
```

The clear separation between Day 0 and Day 7 samples indicates that STAT5B's network neighborhood is substantially rewired upon T-cell activation — despite its expression level remaining unchanged (p = 0.51, T-test on pseudobulk expression).

## Session Info

```{r session-info}
sessionInfo()
```
