---
title: "Unsupervised Sample Subgroup Detection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Unsupervised Sample Subgroup Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

This vignette demonstrates MOSAIC's **unsupervised sample subgroup detection** pipeline. Unlike global approaches that compute patient similarity across all features, MOSAIC groups multi-modal features into coherent connectivity modules and tests whether specific modules delineate distinct patient subtypes. This modular strategy can reveal cryptic subgroup structure that global methods miss.

We apply MOSAIC to a multi-omic prefrontal cortex (PFC) cohort with paired snRNA-seq and snATAC-seq from HIV+ individuals, focusing on L2/3 inhibitory neurons.

## Dataset

We use a single-nucleus multi-omic dataset of human prefrontal cortex tissue comprising 31 donors (18 HIV+, 13 controls) with paired snRNA-seq and snATAC-seq ([Farris et al., 2024](https://doi.org/10.1038/s41467-024-53624-8)). We focus on **L2/3-IT inhibitory neurons** and restrict the analysis to HIV samples only (excluding controls), treating disease labels as unknown.

## Data Preprocessing

```{r load-data}
library(MOSAIC)
library(Seurat)
library(ggplot2)
library(dplyr)
library(dynamicTreeCut)
library(uwot)
library(pheatmap)

# Load preprocessed Seurat objects
rna_seurat <- readRDS("path/to/PFC_L23-IT_rna.RDS")
atac_seurat <- readRDS("path/to/PFC_L23-IT_atac.RDS")

# Use gene activity scores for ATAC
DefaultAssay(atac_seurat) <- "GeneACT"
atac_seurat[["ATAC"]] <- NULL
```

### Sample and Feature Filtering

```{r filtering}
# Remove samples with too few cells
sample_counts <- table(rna_seurat$sample_name)
samples_to_remove <- names(sample_counts[sample_counts < 10])
rna_seurat <- subset(rna_seurat, subset = sample_name %ni% samples_to_remove)
atac_seurat <- subset(atac_seurat, subset = sample_name %ni% samples_to_remove)

# Exclude controls — treat HIV subgroup labels as unknown
rna_seurat <- subset(rna_seurat, subset = condition_new %ni% c("CTR"))
atac_seurat <- subset(atac_seurat, subset = condition_new %ni% c("CTR"))

# RNA feature selection: expressed in all samples + variable
counts <- GetAssayData(rna_seurat, assay = "RNA", slot = "counts")
sample_ids <- unique(as.character(rna_seurat$sample_name))
genes_per_sample <- lapply(sample_ids, function(id) {
  cells <- which(rna_seurat$sample_name == id)
  rownames(counts)[rowSums(counts[, cells, drop = FALSE]) > 3]
})
genes_to_keep <- Reduce(intersect, genes_per_sample)
rna_seurat <- FindVariableFeatures(rna_seurat, nfeatures = 10000)
genes_to_keep <- genes_to_keep[genes_to_keep %in% VariableFeatures(rna_seurat)]

# ATAC feature selection
counts_atac <- GetAssayData(atac_seurat, assay = "GeneACT", slot = "counts")
peaks_per_sample <- lapply(sample_ids, function(id) {
  cells <- which(atac_seurat$sample_name == id)
  rownames(counts_atac)[rowSums(counts_atac[, cells, drop = FALSE]) > 5]
})
atac_seurat <- FindVariableFeatures(atac_seurat, nfeatures = 3000)
peaks_to_keep <- Reduce(intersect, peaks_per_sample)
peaks_to_keep <- peaks_to_keep[peaks_to_keep %in% VariableFeatures(atac_seurat)]

# Subset and normalize
seurat_rna <- subset(rna_seurat, features = genes_to_keep)
seurat_atac <- subset(atac_seurat, features = peaks_to_keep)
seurat_rna <- NormalizeData(seurat_rna)
seurat_atac <- NormalizeData(seurat_atac)

# Feature name vector
feature_names <- c(
  rownames(seurat_rna[["RNA"]]),
  paste0("ATAC-", rownames(seurat_atac[["GeneACT"]]))
)
```

## Step 1: Run MOSAIC Embedding

```{r run-mosaic}
result <- run_MOSAIC(
  seurat_list = list(RNA = seurat_rna, ATAC = seurat_atac),
  assays = c("RNA", "GeneACT"),
  sample_meta = "sample_name",
  condition_meta = "condition_new"
)
```

## Step 2: Compute Per-Feature Stratification Profiles

For each feature, we extract its per-sample similarity matrix (how similarly does this feature behave across patients). We then compute a feature-feature correlation matrix based on these stratification profiles — features that stratify patients in similar ways are grouped into modules.

```{r stratification-profiles}
# Get per-feature similarity matrices (without running PERMANOVA)
n_sample <- length(result$mosaic_embed_list)
dc_result <- run_DC_test(
  result$mosaic_embed_list,
  n_sample = n_sample,
  groups = result$annotation$Condition,
  dist_method = "cosine"
)

# Extract upper triangle of each similarity matrix as a stratification vector
sim_vectors <- lapply(dc_result$similarity_matrix_list, function(mat) {
  mat[upper.tri(mat, diag = FALSE)]
})

# Feature-feature Spearman correlation
strat_matrix <- do.call(cbind, sim_vectors)
feature_correlation_matrix <- cor(strat_matrix, method = "spearman")
```

## Step 3: Identify Feature Modules

We apply hierarchical clustering with dynamic tree cut to identify coherent groups of features that co-vary in their patient stratification patterns.

```{r module-identification}
dist_matrix <- as.dist(1 - feature_correlation_matrix)
hclust_features <- hclust(dist_matrix, method = "ward.D2")

feature_clusters <- cutreeDynamic(
  dendro = hclust_features,
  distM = as.matrix(dist_matrix),
  minClusterSize = 20,
  method = "hybrid",
  deepSplit = 2
)
names(feature_clusters) <- colnames(feature_correlation_matrix)

module_ids <- sort(unique(feature_clusters[feature_clusters > 0]))
cat(length(module_ids), "feature modules identified\n")
```

### Feature Embedding UMAP Colored by Module

We visualize the feature-level embedding to confirm that modules form coherent clusters.

```{r feature-umap, fig.width = 8, fig.height = 5.5}
# PCA on the feature correlation matrix, then UMAP
pca_res <- prcomp(feature_correlation_matrix, scale. = TRUE)
n_pcs <- find_elbow_kneedle(pca_res$sdev[1:100])
umap_res <- umap(pca_res$x[, 1:n_pcs], n_neighbors = 30, min_dist = 0.3)

# Color palette for modules
pal36 <- c(
  "#4E79A7","#A0CBE8","#F28E2B","#FFBE7D","#59A14F","#8CD17D",
  "#B6992D","#F1CE63","#499894","#86BCB6","#E15759","#FF9D9A",
  "#79706E","#BAB0AC","#D37295","#FABFD2","#B07AA1","#D4A6C8",
  "#9D7660","#D7B5A6","#2E91E5","#1CA71C","#FB0D0D","#DA16FF",
  "#222A2A","#B68100","#EB663B","#750D86","#511CFB","#00A08B",
  "#FB00D1","#FC0080","#B2828D","#6C7C32","#778AAE","#862A16"
)
cluster_ids <- sort(unique(feature_clusters))
cluster_colors <- setNames(pal36[seq_along(cluster_ids)], cluster_ids)

df_umap <- data.frame(
  UMAP1 = umap_res[, 1],
  UMAP2 = umap_res[, 2],
  module = as.factor(feature_clusters)
)

ggplot(df_umap, aes(UMAP1, UMAP2, color = module)) +
  geom_point(size = 0.7) +
  scale_color_manual(values = cluster_colors, name = "Feature Module") +
  theme_classic(base_size = 16) +
  ggtitle("Feature Embedding Colored by Module") +
  theme(legend.key.size = unit(0.4, "cm"))
```

## Step 4: Screen Modules for Patient Subgroups

For each module, we compute a module-specific sample similarity matrix using `compute_module_similarity()`, then test whether it reveals a balanced two-way partition with `find_partition_hclust()`. Significance is assessed by a permutation test comparing the observed silhouette score against random feature sets of the same size.

```{r screen-modules}
N_PERM <- 1000
total_features <- nrow(result$mosaic_embed_list[[1]])

module_results <- data.frame(
  module = integer(), n_features = integer(),
  silhouette = numeric(), perm_pval = numeric(),
  group1_size = integer(), group2_size = integer(),
  stringsAsFactors = FALSE
)
module_assignments <- list()
module_sim_matrices <- list()
module_hclust_list <- list()

for (m in module_ids) {
  chosen_idx <- which(feature_clusters == m)
  n_feat <- length(chosen_idx)

  sim_mat <- compute_module_similarity(result$mosaic_embed_list, chosen_idx)
  dist_mat <- as.dist(1 - sim_mat)
  partition <- find_partition_hclust(dist_mat)

  if (!partition$balanced) {
    module_results <- rbind(module_results, data.frame(
      module = m, n_features = n_feat,
      silhouette = NA, perm_pval = NA,
      group1_size = NA, group2_size = NA
    ))
    next
  }

  sil_score <- partition$silhouette

  # Permutation test
  null_sils <- numeric(N_PERM)
  for (p in seq_len(N_PERM)) {
    rand_idx <- sample(total_features, n_feat, replace = FALSE)
    rand_sim <- compute_module_similarity(result$mosaic_embed_list, rand_idx)
    rand_dist <- as.dist(1 - rand_sim)
    rand_part <- find_partition_hclust(rand_dist)
    null_sils[p] <- ifelse(rand_part$balanced, rand_part$silhouette, 0)
  }
  perm_pval <- (sum(null_sils >= sil_score) + 1) / (N_PERM + 1)

  group_sizes <- table(partition$groups)
  module_results <- rbind(module_results, data.frame(
    module = m, n_features = n_feat,
    silhouette = sil_score, perm_pval = perm_pval,
    group1_size = group_sizes[1], group2_size = group_sizes[2]
  ))
  module_assignments[[as.character(m)]] <- partition$groups
  module_sim_matrices[[as.character(m)]] <- sim_mat
  module_hclust_list[[as.character(m)]] <- partition$hclust
}

# Multiple testing correction
module_results$padj <- p.adjust(module_results$perm_pval, method = "BH")
module_results <- module_results[order(module_results$padj, -module_results$silhouette), ]

sig_modules <- module_results[!is.na(module_results$padj) & module_results$padj < 0.05, ]
cat(nrow(sig_modules), "significant modules found\n")
print(sig_modules)
```

## Step 5: Visualize a Significant Module

We select the top significant module and visualize its sample similarity heatmap, annotated by the discovered subgroups and known conditions.

```{r module-heatmap, fig.width = 6.2, fig.height = 5.5}
# Pick the top module (most balanced + highest silhouette)
MODULE_ID <- sig_modules$module[1]
m_char <- as.character(MODULE_ID)

sim_mat <- module_sim_matrices[[m_char]]
groups <- module_assignments[[m_char]]
hc <- module_hclust_list[[m_char]]

# Build annotation
annotation <- result$annotation
annotation$Subgroup <- ifelse(groups == 1, "Group1", "Group2")

annotation_colors <- list(
  Subgroup = c("Group1" = "#4DBBD5", "Group2" = "#E64B35")
)
if ("Condition" %in% colnames(annotation)) {
  conditions <- unique(annotation$Condition)
  cond_palette <- c("HIV" = "#FC8D62", "HIVOUD" = "#E5C494",
                    "OUD" = "#8DA0CB", "CTR" = "#66C2A5")
  annotation_colors$Condition <- cond_palette[intersect(conditions, names(cond_palette))]
}

m_row <- module_results[module_results$module == MODULE_ID, ]

pheatmap(
  sim_mat,
  annotation_row = annotation,
  annotation_col = annotation,
  annotation_colors = annotation_colors,
  cluster_rows = hc,
  cluster_cols = hc,
  show_rownames = FALSE,
  show_colnames = FALSE,
  main = sprintf("Module %d | %d features | Silhouette = %.3f",
                 MODULE_ID, m_row$n_features, m_row$silhouette)
)

# Cross-tabulation
cat("\nSubgroup x Condition:\n")
print(table(annotation$Subgroup, annotation$Condition))
```

## Step 6: Differential Analysis Between Subgroups

We assign cells to the discovered subgroups based on their donor's group membership, then perform differential expression analysis between the two subgroups to characterize their molecular differences.

```{r differential-analysis, fig.width = 6, fig.height = 5}
# Assign subgroup labels to cells
group1_samples <- names(groups[groups == 1])
group2_samples <- names(groups[groups == 2])

rna_seurat$subgroup <- ifelse(
  rna_seurat$sample_name %in% group1_samples, "Group1",
  ifelse(rna_seurat$sample_name %in% group2_samples, "Group2", NA)
)
rna_seurat <- subset(rna_seurat, subset = !is.na(subgroup))

# RNA differential expression
Idents(rna_seurat) <- "subgroup"
de_rna <- FindMarkers(rna_seurat,
                       ident.1 = "Group1", ident.2 = "Group2",
                       test.use = "MAST",
                       logfc.threshold = 0.25, min.pct = 0.1)

de_rna$gene <- rownames(de_rna)
de_rna$significance <- "Not Significant"
de_rna$significance[de_rna$p_val_adj < 0.01 & de_rna$avg_log2FC > 1] <- "Up in Group1"
de_rna$significance[de_rna$p_val_adj < 0.01 & de_rna$avg_log2FC < -1] <- "Down in Group1"

cat(sprintf("RNA DE: %d up, %d down (padj<0.01, |log2FC|>1)\n",
            sum(de_rna$significance == "Up in Group1"),
            sum(de_rna$significance == "Down in Group1")))

# Volcano plot
ggplot(de_rna, aes(x = avg_log2FC, y = -log10(p_val_adj), color = significance)) +
  geom_point(alpha = 0.6, size = 1.5) +
  scale_color_manual(values = c("Up in Group1" = "#D73027",
                                "Down in Group1" = "#4575B4",
                                "Not Significant" = "grey")) +
  geom_hline(yintercept = -log10(0.01), linetype = "dashed") +
  geom_vline(xintercept = c(-1, 1), linetype = "dashed") +
  theme_minimal() +
  labs(title = sprintf("RNA DE Between Module %d Subgroups", MODULE_ID),
       x = "Log2 Fold Change", y = "-Log10 Adjusted P-value")
```

### Pathway Enrichment

```{r pathway-enrichment}
library(clusterProfiler)
library(ReactomePA)
library(org.Hs.eg.db)

up_genes <- de_rna$gene[de_rna$significance == "Up in Group1"]
if (length(up_genes) > 5) {
  entrez_up <- bitr(up_genes, fromType = "SYMBOL",
                    toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  reactome_up <- enrichPathway(gene = entrez_up$ENTREZID,
                                pvalueCutoff = 0.1, readable = TRUE)
  if (nrow(as.data.frame(reactome_up)) > 0) {
    print(dotplot(reactome_up, showCategory = 15, font.size = 8) +
            ggtitle("Pathways Enriched in Group1"))
  }
}

down_genes <- de_rna$gene[de_rna$significance == "Down in Group1"]
if (length(down_genes) > 5) {
  entrez_down <- bitr(down_genes, fromType = "SYMBOL",
                      toType = "ENTREZID", OrgDb = org.Hs.eg.db)
  reactome_down <- enrichPathway(gene = entrez_down$ENTREZID,
                                  pvalueCutoff = 0.1, readable = TRUE)
  if (nrow(as.data.frame(reactome_down)) > 0) {
    print(dotplot(reactome_down, showCategory = 15, font.size = 8) +
            ggtitle("Pathways Enriched in Group2"))
  }
}
```

## Session Info

```{r session-info}
sessionInfo()
```
